Hierarchically Supervised Latent Dirichlet Allocation

### 2011-10-27 paper/experiments.tex
Added poster and updated paper for camera-ready

Discharge summaries are authored by clinicians to summarize patient
hospitalization course. The summaries typically contain a record of patient
complaints, findings and diagnoses, along with treatment and hospital course.
For each hospitalization, trained medical coders review the information in the
discharge summary and assign a series of diagnoses codes. Coding follows the
ICD-9-CM controlled terminology, an international diagnostic classification for
epidemiological, health management, and clinical
purposes.\footnote{http://www.cdc.gov/nchs/icd/icd9cm.htm}  The ICD-9 codes are organized in a rooted-tree structure, with each
edge representing an is-a relationship between parent and child, such that the
parent diagnosis subsumes the child diagnosis. For example, the code for
{}``Pneumonia due to adenovirus'' is a child of the code for {}``Viral
pneumonia,'' where the former is a type of the latter.  It is worth noting that
the coding can be noisy. Human coders sometimes disagree~\cite{Challenge07},
tend to be more specific than sensitive in their
assignments~\cite{Birmetal2005}, and sometimes make
mistakes~\cite{FarzandipourEtAl10}.

@@

Our dataset was gathered from the clinical data warehouse of
a large metropolitan hospital. It consists of 6,000 discharge summaries and
their associated ICD-9 codes (7,298 distinct codes overall), representing all
the discharges from the hospital in 2009. Summaries have 8.39 associated ICD-9
codes on average (std dev=5.01) and contain an average of 536.57 terms after
preprocessing (std dev=300.29). We split our dataset into 5,000 discharge
summaries for training and 1,000 for testing.

@@

sLDA with independent regressors is the most salient comparison model
for our work. The distinguishing factor between HSLDA and sLDA is the
addition structure imposed on the label space, a distinction that we
hypothesized would result in a difference in predictive performance.

@@

\begin{figure}[h]
\begin{center}
%\subfloat[][\label{fig:1a}]{\includegraphics[width=.49\textwidth]{figs/amazon_pred_varying_mu}}
\subfigure[][]{\label{fig:1a}\includegraphics[width=.49\textwidth]{figs/clin_pred_varying_mu}}
%\subfloat[][\label{fig:1c}]{\includegraphics[width=.49\textwidth]{figs/sens_comparison_leafs}}
\subfigure[][]{\label{fig:1b}\includegraphics[width=.49\textwidth]{figs/amazon_pred_varying_mu}}
\subfigure[][]{\label{fig:clinical_roc}\includegraphics[scale=0.4]{figs/ROC_comparison_leafs}}
\caption{ROC curves for out-of-sample ICD-9 code prediction from patient free-text discharge records 
(\subref{fig:1a},\subref{fig:clinical_roc}). ROC curve for out-of-sample Amazon product category predictions from 
product free-text descriptions \subref{fig:1b}. Figures \subref{fig:1a} and \subref{fig:1b} are a function 
of the prior means of the regression parameters. Figure \subref{fig:clinical_roc} is a function of auxiliary variable threshold. In all figures, solid is 
HSLDA, dashed are independent regressors + sLDA (hierarchical 
constraints on labels ignored), and dotted is HSLDA fit by running LDA first then running 
tree-conditional regressions.}
\label{fig:main_results}
\end{center}
\end{figure}

@@

We evaluated our model, HSLDA, against the comparison models with a focus on
predictive performance on held-out data. Prediction performance was measured
with standard metrics -- sensitiviy (true positive rate) and 1-specificity
(false positive rate).

The gold standard for comparison was derived from the testing set in each
dataset. To make the comparison as fair as possible among models, ancestors of
observed nodes in the label hierarchy were ignored, observed nodes were
considered positive and descendents of observed nodes were considered to be
negative. In particular, since the sLDA model does not enforce the hierarchical
constraints, its performance was found to be so poor on joint prediction that
it was not included. The models can be compared on a more equal footing by
considering only on the observed labels as being positive, despite the fact
that, following the hierarchical constraints, ancestors must also be positive.
Such a gold standard will likely inflate the number of false positives because
the labels applied to any particular document are usually not as complete as
they could be.  ICD-9 codes, for instance, lack sensitivity and their use as a
gold standard could lead to correctly positive predictions being labeled as
false positives\citep{Birmetal2005}. 
However, given that the label space is often large (as in our examples) it is a
moderate assumption that erroneous false positives should not skew results
significantly.

Predictive performance in HSLDA is evaluated by $p\left(y_{l,\hat{d}}\mid w_{1:N_{\hat{d}},\hat{d}}, w_{1:N_d,1:D},  y_{l\in\mathcal{L},1:D}\right)$ 
for each test document, $\hat{d}$. For efficiency, the expectation of this
probability distribution was estimated in the following way. Expectations 
of $\mathbf{\bar{z}}_d$ and $\boldsymbol{\eta}_l$ were estimated with samples
from the posterior. Using these expectations, we performed Gibbs sampling over
the hierarchy to acquire predictive samples for the documents in the test set.
The true positive rate was calculated as the average expected labeling for
gold standard positive labels. The false positive rate was calculated 
as the average expected labeling for gold standard negative labels.

Figure~\ref{fig:1a} shows the results for the clinical data. For instance, 
a prior mean for the regression of $\mu=-1.6$ yields the following performance:
the full HSLDA model had a 
true positive rate of 0.57 and a false positive rate of 0.13, the sLDA model
had a true positive rate of 0.42 and a false positive rate of 0.07, and the
HSLDA model where LDA and the regressions were fit separately had a true
positive rate of 0.39 and a false positive rate of 0.08.

@@

Figure~\ref{fig:1b} shows the results for the retail product data. For
instance, a prior mean for the regression of $\mu=-2.2$ yields the following
performance: the full HSLDA model had a true positive rate of 0.85 
and a false positive rate of 0.30, the sLDA model had a true positive
rate of 0.78 and a false positive rate of 0.14, and the HSLDA model where
LDA and the regressions were fit separately had a true positive rate of 0.77
and a false positive rate of 0.16. These results follow a similar pattern to the clinical data.

Figure \ref{fig:main_results} shows the predictive performance of HSLDA 
relative to the two comparison models as a function of the prior mean
on regression coefficients as a receiver operating characteristic (ROC) curve. 
For low values of the auxiliary variable threshold, the models predict labels
in a more sensitive and less specific manner, creating the points in the upper
right corner of the ROC curve. As the auxiliary variable threshold is
increased, the models predict in a less sensitive and more specific manner,
creating the points in the lower left hand corner of the ROC curve. For all
values of the prior mean in both datasets, HSLDA outperforms sLDA with
independent regressors. In the case of HSLDA with separately trained
regression, HSLDA outperforms in the clinical dataset but performs equally well
across the board with the retail product dataset.

>>>>

Discharge summaries are authored by clinicians to summarize patient
hospitalization courses. The summaries typically contain a record of patient
complaints, findings and diagnoses, along with treatment and hospital course.
For each hospitalization, trained medical coders review the information in the
discharge summary and assign a series of diagnoses codes. Coding follows the
ICD-9-CM controlled terminology, an international diagnostic classification for
epidemiological, health management, and clinical
purposes.\footnote{http://www.cdc.gov/nchs/icd/icd9cm.htm}  The ICD-9 codes are organized in a rooted-tree structure, with each
edge representing an is-a relationship between parent and child, such that the
parent diagnosis subsumes the child diagnosis. For example, the code for
{}``Pneumonia due to adenovirus'' is a child of the code for {}``Viral
pneumonia,'' where the former is a type of the latter.  It is worth noting that
the coding can be noisy. Human coders sometimes disagree~\cite{Challenge07},
tend to be more specific than sensitive in their
assignments~\cite{Birmetal2005}, and sometimes make
mistakes~\cite{FarzandipourEtAl10}.

@@

Our dataset was gathered from the clinical data warehouse of
a large metropolitan hospital. It consists of 6,000 discharge summaries and
their associated ICD-9 codes (7,298 distinct codes overall), representing all
the discharges from the hospital in 2009. All included discharge summaries had associated ICD-9 Codes.
Summaries have 8.39 associated ICD-9
codes on average (std dev=5.01) and contain an average of 536.57 terms after
preprocessing (std dev=300.29). We split our dataset into 5,000 discharge
summaries for training and 1,000 for testing.

@@

sLDA with independent regressors is the most salient comparison model
for our work. The distinguishing factor between HSLDA and sLDA is the
additional structure imposed on the label space, a distinction that we
hypothesized would result in a difference in predictive performance.

@@

\begin{figure}[h]
\begin{center}
%\subfloat[][\label{fig:1a}]{\includegraphics[width=.49\textwidth]{figs/amazon_pred_varying_mu}}
\subfigure[][Clinical data performance.]{\label{fig:1a}\includegraphics[width=.49\textwidth]{figs/clin_pred_varying_mu}}
%\subfloat[][\label{fig:1c}]{\includegraphics[width=.49\textwidth]{figs/sens_comparison_leafs}}
\subfigure[][Retail product performance.]{\label{fig:1b}\includegraphics[width=.49\textwidth]{figs/amazon_pred_varying_mu}}
%\subfigure[][]{\label{fig:clinical_roc}\includegraphics[scale=0.4]{figs/ROC_comparison_leafs}}
\caption{ROC curves for out-of-sample label prediction varying $\mu$, the prior mean of the regression parameters. 
In both figures, solid is HSLDA, dashed are independent regressors + sLDA (hierarchical 
constraints on labels ignored), and dotted is HSLDA fit by running LDA first then running 
tree-conditional regressions.}
%\label{fig:main_results}
\end{center}
\end{figure}

@@

We evaluated our model, HSLDA, against the comparison models with a focus on
predictive performance on held-out data. Prediction performance was measured
with standard metrics -- sensitivity (true positive rate) and 1-specificity
(false positive rate).

The gold standard for comparison was derived from the testing set in each
dataset. To make the comparison as fair as possible among models, ancestors of
observed nodes in the label hierarchy were ignored, observed nodes were
considered positive and descendents of observed nodes were considered to be
negative. In particular, since the sLDA model does not enforce the hierarchical
constraints, we establish a more equal footing by
considering only the observed labels as being positive, despite the fact
that, following the hierarchical constraints, ancestors must also be positive.
Such a gold standard will likely inflate the number of false positives because
the labels applied to any particular document are usually not as complete as
they could be.  ICD-9 codes, for instance, lack sensitivity and their use as a
gold standard could lead to correctly positive predictions being labeled as
false positives~\citep{Birmetal2005}. 
However, given that the label space is often large (as in our examples) it is a
moderate assumption that erroneous false positives should not skew results
significantly.

Predictive performance in HSLDA is evaluated by $p\left(y_{l,\hat{d}}\mid w_{1:N_{\hat{d}},\hat{d}}, w_{1:N_d,1:D},  y_{l\in\mathcal{L},1:D}\right)$ 
for each test document, $\hat{d}$. For efficiency, the expectation of this
probability distribution was estimated in the following way. Expectations 
of $\mathbf{\bar{z}}_{\hat{d}}$ and $\boldsymbol{\eta}_l$ were estimated with samples
from the posterior. Using these expectations, we performed Gibbs sampling over
the hierarchy to acquire predictive samples for the documents in the test set.
The true positive rate was calculated as the average expected labeling for
gold standard positive labels. The false positive rate was calculated 
as the average expected labeling for gold standard negative labels.

\begin{figure}[t]
%tbp] %  figure placement: here, top, bottom, or page
\centering \includegraphics[scale=0.4]{figs/ROC_comparison_leafs} \caption{ROC curve for out-of-sample ICD-9 code prediction varying auxiliary variable threshold. $\mu = -1.0$ for all three models in this figure.}
\label{fig:results_auxiliary_var} 
\end{figure}

Figure~\ref{fig:1a} demonstrates the performance of the model on the clinical data as an ROC curve varying $\mu$.  For instance, 
a hyperparameter setting of $\mu=-1.6$ yields the following performance:
the full HSLDA model had a 
true positive rate of 0.57 and a false positive rate of 0.13, the sLDA model
had a true positive rate of 0.42 and a false positive rate of 0.07, and the
HSLDA model where LDA and the regressions were fit separately had a true
positive rate of 0.39 and a false positive rate of 0.08. These points are highlighted in Figure~\ref{fig:1a}.

@@

Figure~\ref{fig:1b} demonstrates the performance of the model on the retail product data as an ROC curve also varying $\mu$. For
instance, a hyperparameter setting of $\mu=-2.2$ yields the following
performance: the full HSLDA model had a true positive rate of 0.85 
and a false positive rate of 0.30, the sLDA model had a true positive
rate of 0.78 and a false positive rate of 0.14, and the HSLDA model where
LDA and the regressions were fit separately had a true positive rate of 0.77
and a false positive rate of 0.16. These results follow a similar pattern to the clinical data. These points are highlighted in Figure~\ref{fig:1b}.

Figure \ref{fig:results_auxiliary_var} shows the predictive performance of HSLDA 
relative to the two comparison models on the clinical dataset as a function of the auxiliary variable threshold. 
For low values of the auxiliary variable threshold, the models predict labels
in a more sensitive and less specific manner, creating the points in the upper
right corner of the ROC curve. As the auxiliary variable threshold is
increased, the models predict in a less sensitive and more specific manner,
creating the points in the lower left hand corner of the ROC curve. HSLDA with full joint inference outperforms sLDA with
independent regressors as well as HSLDA with separately trained
regression.
